\documentclass{ciit-dissertation}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{caption}
\usepackage{verbatim}
\usepackage{fancyvrb}
\usepackage{graphicx}


\addbibresource{thesis.bib}


% Declare all of the data required to create the frontmatter (title pages, declarations .etc)
\title{Beam Forming With Phased Array Antennas}
\author{Laaraib}
\registrationnumber{CIIT/SP14-BPH-024/ISB}
\department{Physics}
\campus{Islamabad}
\degreetype{\undergraduate}
\program{BS Physics}
\dissertationtype{Thesis}
\session{Fall 2017}
\submissiondate{Jan 2, 2018}

\supervisor{Dr. Abid H. Mujtaba}
\supervisordesignation{Assistant Professor}

\hod{Dr. Sajid Qamar}
\hoddesignation{Professor}

\externalexaminer{}
\externalexaminerdesignation{}
\externalexaminerinstitution{}
\externalexamineraddress{Islamabad}

\dedication{I Dedicate My Thesis To My Dear Parents, Mr. and Mrs. Muhammad Mansha.}

\acknowledgement{After all the struggles and hard work of a complete year, now is the time that I pay my gratitude to the people responsible for making this possible for me.
	
First and foremost, I owe a great debt of gratitude to My supervisor "Dr. Abid Mujtaba", for the continuous support of My project work and related research, for his patience, motivation and immense knowledge. Without his assistance and dedicated involvement in every step throughout the process, this thesis would have never been accomplished. Thank you sir, for your sage advices and vital guidance, which made it possible to complete My project and writing of this thesis.

Most importantly, I am deeply grateful to My Parents and Siblings, for their attention, support and being there for me throughout My life.
		
		Thank you!}

\abstract{%
To create a high gain antenna, which radiates radio waves in a narrow beam, two general techniques can be used. One technique is to use large metal surfaces such as parabolic reflectors, horns or dielectric lenses which change the direction of the radio waves by reflection or refraction, to focus the radio waves from a single low gain antenna into a beam. This type is called an aperture antenna.

The second technique is to use multiple antennas to form a directed beam (beam forming) which involves the operations of phase shifting and amplitude tapering for each element; this is called a phased array antenna, or antenna array. Depending upon their geometry we have different configurations of arrays for example linear, circular and complex array.
	
This project used the second technique to derive, calculate and simulate the directed beam (one with maximum intensity at a location of my choice) created by multiple antenna sources arranged in a linear array with the freedom to manage only the phase between them.
}

\begin{document}
	
	\makefrontmatter        % Create all of the pages before the first chapter: title, dedication, ack, abstract, ToC, LoF .etc
	
    \include{chap1}
    \include{chap2}
    \include{chap3}
    \include{chap4}
    \include{chap5}
	
    \begin{appendices}
      \include{code}
    \end{appendices}
	
	\makereferences         % Create the references section
	
\end{document}


